### 2015-05-01 papers/histogram/thesis/thesis.tex
additions to histogram thesis

\renewcommand{\title}{Optimizing Monte Carlo state space exploration
  of the square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{21 April 2015}

@@

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased (i.e. Metropolis) Monte Carlo
simulations are incapable of sampling the low energy fluid states
which dominate the partition function at low (i.e. liquid-state)
temperatures. Previous works have developed several generic Monte
Carlo histogram methods for collecting statistics on low energy system
states, but little or no literature exists on these methods'
systematic comparison, as well as their application to the square-well
fluid. The square-well fluid in particular introduces application
challenges not manifest in traditional models for testing and
benchmarking such numerical techniques (e.g. the Ising model). We
propose to implement these methods in Monte Carlo simulations of the
square-well fluid, determine appropriate performance metrics for their
comparison, and identify those methods which are most efficient and
effective at sampling otherwise improbable system states.

@@

  \item Attempt to change the position of one sphere (in general, a
    single fluid ``atom'' or ``molecule'') to a random location,
    rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:mc_move}

@@

  \item Attempt to change the position of one sphere, rejecting the
    change if
    \begin{enumerate*}[label=\roman*)]
    \item it results in a forbidden fluid configuration, or
    \item a newly chosen random number on the interval $\sp{0,1}$ is
      larger than the probability given by (\ref{eq:move_prob}).
    \end{enumerate*}
    \label{alg:biased_mc_move}

@@

The problem with this implementation of the flat histogram method is
that histogram methods are generally necessary when some range of
energies $R$ is inaccessible via Metropolis Monte Carlo
simulations. In such a case, the energy histogram $H\p{E\in R}$ after
simulating for any reasonable amount of time will be statistically
insignificant or null, invalidating the approximation
$H\p{E}\approx\tilde D\p{E}$. This method will therefore only
``flatten'' the histogram at energies $E$ which have been sufficiently
sampled for $H\p{E}$ to be statistically significant. The end goal in
implementing histogram methods, however, is precisely to sample those
energies that cannot be sufficiently sampled without employing clever
schemes to bias Monte Carlo simulations. The flat histogram method
therefore does not directly help us study the square-well fluid, but
it does provide the basis for the method in Section
\ref{sec:simple_flat}.

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_analysis, wang_landau_mod}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations. \fixme{more intro
  on WL}

@@

Interestingly, the flatness condition $C\sp{H}$ for updating $f$ and
restarting the process should not be too stringent: if the histogram
is completely flat, say $H\p{E}=H_0$ for all $E$, then the weights
$w\p{E}$ have all been modified by the same factor of $f^{H_0}$.
Remembering the scale-invariance of $w\p{E}$, this result would mean
that the weights have not actually changed at all! Similarly, too lax
of a flatness condition will prompt the initialization process to
\begin{enumerate*}[label=\roman*)]
\item reset and decrease $f$ too early, i.e. when the weight array
  could still use some heavy handed adjustment, and
\item quit before one has any reason to think that simulations might
  yield a flat histogram.
\end{enumerate*}
Tuning the flatness condition is thus crucial to the success of the
Wang-Landau method.

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
initialization process, which has five free parameters: the number of
iterations $n$ to run in step \ref{alg:wang_landau_sim}, the initial
factor $f_0$, the update factor $u$, the flatness condition $C\sp{H}$,
and the cutoff $c$. The first of these parameters, $n$, controls how
often to check the flatness condition $C\sp{H}$ and run updates, if
necessary. The value of $n$ is not too important, so long as it is
reasonable. Too small value of $n$ will waste computation time
determining $C\sp{H}$, while too large of a value will waste time
simulating when $C\sp{H}$ has already been satisfied. In general, $n$
should be proportional to the computation time of $C\sp{H}$, which
scales with energy range of the simulated system. We therefore use
$n=L$, where $L$ is the number of energy levels for the given
square-well fluid (as discussed in Section \ref{sec:simple_flat}).
Papers in the literature typically neglect providing any heuristic or
formula for $n$, reporting a system-independent $n=10^3$ or
$10^5$\cite{wang_landau, wang_landau_mod}.

The next parameter, $f_0$, controls the initial amount by which to
adjust the weight of the current energy after each move. As mentioned
in Algorithm \ref{alg:wang_landau}, one should always have $f_0>1$,
but the appropriate value of $f_0$ will generally depend on the system
at hand. Too large a value of $f_0$ will magnify stochastic error in
initialization, while too small a value of $f_0$ will cause
simulations to take an exceedingly long time to converge.  Lacking any
heuristics for assigning $f_0$, literature typically suggests
$f_0=e$\cite{wang_landau}; after some experimentation with the free
parameters in the Wang-Landau method, we used $f_0=e^{1/8}$ (in
practice, one usually works with $\ln f$, making the commonly
suggested value $\ln f_0=1$ and our own $\ln f_0=1/8$).

The factor $u$ appearing in step \ref{alg:wang_landau_update} of
Algorithm \ref{alg:wang_landau} controls the amount by which to adjust
$f$ when the histogram $H\p{E}$ is sufficiently flat. In principle,
one need not take a root of $f$ at each reset: the general idea behind
Algorithm \ref{alg:wang_landau} requires only $f_k<f_{k-1}$ to
converge (assuming that $C\sp{H}$ can be satisfied) and
$\lim_{k\to\infty}f_k=1$ to terminate. Taking the $u$-th root of $f$
at each increment of $k$ is a natural and convenient iterative means
to satisfy these requirements. Previous papers have used $u=2$, which
we have adopted as well\cite{wang_landau, wang_landau_mod}.

\fixme{Walk through $C\sp{H}$ and $c$. Finish discussion of WL.}

\subsection{The optimized ensemble (OE) method}
\label{sec:optimized_ensemble}

\fixme{Explain, provide algorithm.}

\begin{algorithm}[H]
  \caption{Modifying weights via the optimized ensemble}
  \label{alg:optimized_ensemble}
\end{algorithm}

\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

@@

@@

@@

\fixme{This section might just become a part of the section on the
  minimum important energy. (\ref{eq:density_of_states}) gives us an
  explicit formula for the density of states if we are confident in
  our energy histogram. We can also compute the density of states from
  the transition matrix, which is sometimes more appropriate to do.}

>>>>

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid} \renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy} \renewcommand{\date}{21
  April 2015}

@@

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We propose to implement these
methods in Monte Carlo simulations of the square-well fluid, determine
appropriate performance metrics for their comparison, and identify
those methods which are most efficient and effective at sampling
otherwise improbable system states.

@@

  \item Randomly attempt to change the position of one sphere (in
    general, a single fluid ``atom'' or ``molecule'') to another
    location, rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:mc_move}

@@

  \item Randomly attempt to change the position of one sphere,
    rejecting the change if
    \begin{enumerate*}[label=\roman*)]
    \item it results in a forbidden fluid configuration, or
    \item a newly chosen random number on the interval $\sp{0,1}$ is
      larger than the probability determined by (\ref{eq:move_prob})
      for the initial and final states $s_i$ and $s_f$, respectively.
    \end{enumerate*}
    \label{alg:biased_mc_move}

@@

The problem with this implementation of the flat histogram method is
that histogram methods are generally necessary when some range of
energies $R$ is inaccessible via Metropolis Monte Carlo
simulations. In such a case, the energy histogram $H\p{E\in R}$ after
simulating for any reasonable amount of time will be statistically
insignificant or null, invalidating the approximation
$H\p{E}\approx\tilde D\p{E}$. This method will therefore only
``flatten'' the histogram at energies $E$ which have been sufficiently
sampled for $H\p{E}$ to be statistically significant. The end goal in
implementing histogram methods, however, is precisely to sample those
energies that cannot be sufficiently sampled without employing clever
schemes to bias Monte Carlo simulations. The flat histogram method
therefore does not directly help us study the square-well fluid. This
method does, however, motivate the method in Section
\ref{sec:simple_flat}, and provides some of the theory behind the
methods in Sections \ref{sec:tmmc} and \ref{sec:oetmmc}.

@@

\fixme{find more sources to cite for some of the claims made in this
  section}

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_mod, wang_landau_analysis}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations. \fixme{more intro
  on WL}

@@

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
initialization process, which has five free parameters: the number of
iterations $n$ to run in step \ref{alg:wang_landau_sim}, the initial
factor $f_0$, the update factor $u$, the flatness condition $C\sp{H}$,
and the cutoff $c$. The first of these parameters, $n$, controls how
often to check the flatness condition $C\sp{H}$ and run updates, if
necessary. The value of $n$ is not too important, so long as it is
reasonable for the system at hand. Too small value of $n$ will waste
computation time determining $C\sp{H}$, while too large of a value
will waste time simulating when $C\sp{H}$ has already been
satisfied. In general, $n$ should be proportional to the computation
time of $C\sp{H}$, which scales with energy range of the simulated
system. We therefore use $n=L$, where $L$ is, as described in Section
\ref{sec:simple_flat}, the number of energy levels for the given
square-well fluid. Other works typically neglect providing any
heuristic or formula for $n$, reporting a system-independent
$n=10^3$\cite{wang_landau_mod} or $10^5$\cite{wang_landau}.
\fixme{find more citations for this last claim}

The next parameter, $f_0$, controls the initial amount by which to
adjust the weight of the current energy after each move. As mentioned
in Algorithm \ref{alg:wang_landau}, one should always have $f_0>1$,
but the appropriate value of $f_0$ will generally depend on the system
at hand. Too large a value of $f_0$ will magnify stochastic error in
initialization, while too small a value of $f_0$ will cause
simulations to take an exceedingly long time to converge. Lacking any
heuristics for assigning $f_0$, literature typically suggests $f_0=e$
\cite{wang_landau}. After some experimentation with the free
parameters in the Wang-Landau method, we decided on $f_0=e^{1/8}$. In
practice, one usually works with $\ln f$, making the commonly
suggested value $\ln f_0=1$ and our own $\ln f_0=1/8$.

The factor $u$ appearing in step \ref{alg:wang_landau_update} of
Algorithm \ref{alg:wang_landau} controls the amount by which to adjust
$f$ when the histogram $H\p{E}$ is sufficiently flat. In principle,
one need not take a root of $f$ at each reset: the general idea behind
Algorithm \ref{alg:wang_landau} requires only $f_k<f_{k-1}$ to
converge (assuming that $C\sp{H}$ can be satisfied) and
$\lim_{k\to\infty}f_k=1$ to terminate. Taking the $u$-th root of $f$
at each increment of $k$ is a natural and convenient iterative means
to satisfy these requirements. Previous works have reported using
values such as $u=2$ \cite{wang_landau, wang_landau_mod} and $u=10$
\cite{wang_landau_analysis}; we use the former, conservatively
favoring more initialization iterations by decreasing $f$ by a smaller
amount on each reset.

One might guess why the flatness condition $C\sp{H}$ for updating $f$
and restarting the Wang-Landau initialization process should not be
made too lax: doing so would prompt the initialization process to
\begin{enumerate*}[label=\roman*)]
\item reset and decrease $f$ too early, i.e. when the weight array
  could still use some heavy handed adjustment, and
\item quit before one has any reason to think that simulations might
  yield a flat histogram.
\end{enumerate*}
Less obvious, however, is the fact that the flatness condition
$C\sp{H}$ should not be made too stringent: if the histogram is
completely flat, say $H\p{E}=H_0$ for all $E$, then the weights
$w\p{E}$ have all been modified by the same factor of $f^{H_0}$.
Remembering the scale-invariance of $w\p{E}$, this result would mean
that the weights have not actually changed at all!

The success of Wang-Landau thus relies on some degree of leniency in
the flatness condition, which makes tuning $C\sp{H}$ crucial to the
efficient and effective implementation of this method. Some previous
works\cite{wang_landau} suggest that $C\sp{H}$ should be made as
stringent as possible, cautioning only that some simulations might not
satisfy too strict of a flatness condition in any reasonable amount of
time. These works fail to recognize the reliance of Wang-Landau on the
{\it failure} of $C\sp{H}$ to enforce a perfectly flat histogram. A
commonly suggested flatness condition is that the minimum histogram
value be at least some fixed proportion of the mean histogram value,
i.e. $C\sp{H}:\min\sp{H\p{E}}\ge x\bk{H}_E$, and several papers report
using $x=0.95$. We used the same condition with $x=0.25$.

The final free parameter in the Wang-Landau method is the cutoff $c$,
which determines how small $f$ can get before the algorithm stops
modifying the weight array. Too large of a value for $c$ will make
Wang-Landau quit prematurely, whereas too small a value will cause the
algorithm to waste time simulating after the weights $w\p{E}$ have
essentially converged. One might imagine using an alternate end
condition for Wang-Landau, which involves comparing the current
weights $w_k\p{E}$ to those at the end of the previous cycle,
$w_{k-1}\p{E}$, to check whether the initialization process is still
making appreciable modifications to the weights, but such a check
would require both a metric and a free parameter to define what an
appreciable change is. For simplicity, we stuck with the standard end
condition given in Algorithm \ref{alg:wang_landau}. As with the other
free parameters in Wang-Landau, we are not aware of any heuristics for
determining an appropriate value of $c$, but literature cites typical
values of $\ln c=10^{-8}$; we used $\ln c=10^{-6}$.

\subsection{The transition matrix Monte Carlo method}
\label{sec:tmmc}

Unlike the previous histogram methods, the transition matrix Monte
Carlo (TMMC) method does not use weights, and merely compute the
density of states $D\p{E}$ of a system. The density of states can in
turn be used via the flat histogram method (Section
\ref{sec:flat_histogram}) to determine appropriate weights for
subsequent Monte Carlo simulations. TMMC introduces a new object: the
energy transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

The transition matrix $T$ is trivially square and positive-definite,
but it is not symmetric. Given that Metropolis Monte Carlo samples all
of state-space randomly, meaning all possible system states are
equally likely to occur during simulation, a transition $E_i\to E_f$
from an energy with a low state density to one with a high state
density, $D\p{E_i}<D\p{E_j}$, is more probable than the inverse
transition $E_j\to E_i$ simply due to the fact that there are more
states with energy $E_i$ than those with energy $E_j$. Thus
$D\p{E_i}<D\p{E_j}$ implies $P\p{E_i\to E_j}>P\p{E_j\to E_i}$, which
means $T_{ji}>T_{ij}$. Said otherwise, the asymmetry of the transition
matrix is a direct consequence of the second law of thermodynamics:
systems are exceedingly likely to evolve towards macroscopic states
with higher accessible microstate densities.

In principle, computationally determining the transition matrix of a
system is straightforward. One need only simulate a system and collect
a histogram $\tilde T\p{E_i,E_j}$ of the energy transitions after each
move, directly after deciding whether to reject the move based on
whether it resulted in an (in)valid system state. That is, transition
data should be collected (only) on all moves as they would have been
accepted by Metropolis Monte Carlo, i.e. before deciding whether to
reject any given move for any reasons involving weights, or other
biasing schemes.

\vspace{3cm}

\fixme{schemes for initializing to collect transition matrix}

\fixme{normalizing transition matrix}

\fixme{computing density of states via transition matrix}

\fixme{storing transition matrix in the form $T\p{E,\Delta E}$}

\vspace{3cm}

For the square-well fluid, as with many systems, there is a maximum
amount $\Delta_ME$ by which a single move can change the energy of the
system. The transition matrix can therefore be stored in the form
$T\p{E,\Delta E}$

the transition matrix is band-diagonal because . In implementation,
the transition matrix can therefore be stored in the form
$T\p{E,\Delta E}$, making it an $L\times\Delta_ME$ matrix, where $L$
is the number of energy levels in a given system. Given that $L$
scales as $N\Delta_ME$, where $N$ is the number of spheres in the
square-well fluid, the memory required to store the transition matrix
therefore scales as $N\p{\Delta_ME}^2$, an improvement over the $N^2$
growth of a transition matrix in the form $T\p{E_i,E_j}$.

@@

\subsection{The optimized ensemble method}
\label{sec:optimized_ensemble}

\fixme{Explain briefly. No need to provide algorithm.}

@@

\fixme{Explain, provide algorithm. This method is our own synthesis of
  other work.}

@@

\fixme{This section might just become a part of the section on the
  minimum important energy. (\ref{eq:dos}) gives us an explicit
  formula for the density of states if we are confident in our energy
  histogram. We can also compute the density of states from the
  transition matrix, which is sometimes more appropriate to do.}
